---
title: "Cross-Reference a Section"
author: "Andreia Pinto"
date: 2022-03-07
description: "This post shows how to cross-reference sections in a Rmd file."
tags: ["bookdown", "cross-reference"]
categories: ["Rmd basics"]
link-citations: true
ShowToc: true
editPost:
    URL: "https://github.com/AntonioFidalgo/post/blob/main/s-crossrefsection/index.Rmd"
    Text: "Rmd Source Code \U0001f4ce"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(bookdown)
```

Cross-references are references to different elements somewhere in the content. In the case of this post, we will approach the referencing to a section in the document.  

# Necessary Package {.tabset}

Because cross-referencing is not readily available with `knitr`, there is a need to install the package `bookdown`.^[Check out here [How to Install Packages](https://antoniofidalgo.page/discoveR/setup.html#Install_Packages).]  

This package is particularly designed for writing books, but can be used for a R Markdown file. To do so, it requires a change in YAML.  

### From `knitr`

```{r, eval=FALSE}
output:
  XXX_document:
```

### To `bookdown` 

```{r, eval=FALSE}
output:
  bookdown:: XXX_document2:
    number_sections: TRUE
```

The option `number_sections: TRUE` will make sure that our sections are numbered. This is very important, as cross-referencing only works if our sections are numbered (C-R returns the number).

# Labeling {#label-2}

The section to be referenced needs to take a label, which will allow us to call the section. There are a few **rules of labeling** to be considered:  

* Labels must be unique, meaning no two sections can have an equal label;  
* Labels must start with a letter (`a` to `z`);  
* Labels can only contain:  
  * alphanumeric characters (letters and numbers), and  
  * the special characters `:`, `-`, `/`;  
* Empty spaces (`blank`) are not allowed;  
* It is advisable that labels are easy to remember.  

Labels are assigned by adding the code `{#label}`, where `label` respects the previous rules.   

For instance, the labels of sections can look like this:   

`# Introduction {#intro}`  
`## Step One {#step1}`  
`### Complexity of Step One {#complex-step1}`  

# Cross-Referencing

The code reference can be achieved with the code `\@ref(label)`, where `label` is the label chosen for the section being referenced.  
If R fails to find the label used in a cross-reference, computation does not stop. Instead, the output will reflect the mistake by returning `??` where the section number was supposed to appear.  

# Example 

As example, let's consider this document's section called "Labeling".  

### Section Header

`# Labeling {#label-2}`

### C-R code

`In the section \@ref(label-2) of this document, I provide a list of rules for correct labeling.`

### Output

In the section \@ref(label-2)^[Of course, here the cross-referencing will not work because the sections are not numbered.] of this document, I provide a list of rules for correct labeling.

